\documentclass[11pt,a4paper]{article}

\usepackage{acl2015}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{ifthen}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{scalerel}
\usepackage{array}
\usepackage[noend]{algpseudocode}
\usepackage[usenames,dvipsnames]{color}
\usepackage{tikz}
\usepackage[all]{nowidow}
\usetikzlibrary{trees,matrix}
\usepackage{afterpage}
\usepackage{multirow,color}
\input{macros.tex}

\title{Compact, Efficient and Unlimited Capacity:
    Language Modelling with Compressed Suffix Trees}

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\}

\date{}

\begin{document}
\maketitle
\begin{abstract}
  Here goes the abstract.
\end{abstract}

\section{Introduction}

LMs important, critical aspects are speed and scale. One avenue of better LMs using NN; another
better count based LMs to harness more data (e.g., stupid backoff). 

We're the latter, but marrying large scale with KN smoothing (better method). Crux is how to compute
the quantities (counts, occ counts). 

Suffix trees / arrays / tries etc used in the past. Either precomputed (KLM/SRI) or on the fly (slow
LREC paper). But doesn't scale. We propose compressed SA/ST methods, which scale to massive datasets.

Brief outline of how the method works; using one or two suffix trees. Forward/backward search.

Pata on experiments: pplx good; can scale to large n and massive corpora; adequate runtime
performance, around 100x slower than SRILM. Reranking gains? Can we run with INFINITE order?

\section{Suffix-based Indexing}
\label{sec-suffix}
\input{sec-suffix.tex}

\section{Language Modelling}
\label{sec-lm}
\input{sec-lm.tex}

\section{Using \CSTs for KN Computation}
\label{sec-lmsdsl}
\input{sec-lmsdsl.tex}

\section{Experiments}
\label{sec-experiments}
%\input{sec-experiments.tex}








\section{Related Work}

\cite{brants2007large}
\cite{guthrie2010storing}
\cite{brants2006web}
\cite{stolcke2011srilm}
\cite{stolcke2002srilm}
\cite{pauls2011faster}
\cite{heafield2011kenlm}
\cite{kennington2012suffix}
\cite{wood2011sequence}
\cite{chen1996empirical}
\cite{kneser1995improved}
\cite{chen1996empirical}
\cite{navarro2007compressed}
\cite{gog2014theory}

\section{Conclusions}

% include your own bib file like this:
\bibliographystyle{acl}
\bibliography{local}

\end{document}
